Vertex Coloring and Applications

### 2017-05-02 work/paper.tex
Draft for paper

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{titling}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[numbib,nottoc]{tocbibind}
\usepackage{enumitem}
\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}
\setlength\parindent{0pt}
\setlength{\droptitle}{-1.5cm}

@@

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{titling}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[numbib,nottoc]{tocbibind}
\usepackage{enumitem}
\usepackage{multicol}
\setlist[itemize]{topsep=0pt}
\setlist[enumerate]{topsep=0pt}
\setlength\parindent{0pt}
\setlength{\droptitle}{-1.5cm}

@@

\begin{prop}
Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not always produce an optimal solution, i.e. they do not always produce a minimum coloring of a graph.
\end{prop}

We will show this for \emph{heuristic \ref{heuristic-welsh}}. It is trivial to show the same for \emph{heuristic \ref{heuristic-greedy}} as \emph{heuristic \ref{heuristic-welsh}} includes the same steps as \emph{heuristic \ref{heuristic-greedy}}. We will construct a simple graph, \(G = (V, E)\), such that:
%
\begin{align}
|V| &\geq 8 \\
\chi(G) &> \chi^{*}(G)
\end{align}
%
Let \(\chi(G)\) refer to the coloring of \(G\) generated by \emph{heuristic \ref{heuristic-welsh}}. Let \(\chi^{*}(G)\) be the optimal coloring of \(G\). We constructed a simple graph, \(G_1 = (V, E)\), such that \(\Delta(G_1) = 3\) and \(\delta(G_1) = 1\).

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/graph-1.png}
\caption{Uncolored original graph \(G_1\)}
\end{figure}

By applying \emph{heuristic \ref{heuristic-welsh}}, we get the following coloring. The numbers indicate the order the vertices were colored in while applying the heuristic. Any vertex of the same degree got assigned arbitrarily. This results in \(\chi(G_1) = 3\).

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/graph-2.png}
  \caption{Coloring of $G_1$ after applying heuristic \ref{heuristic-welsh}}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/graph-3.png}
  \caption{Optimal coloring of graph \(G_1\)}\label{fig:ex1-optimal}
  \end{figure}
\end{multicols}

\begin{definition}[Bipartite graph]
A bipartite graph is one whose vertex set can be partitioned into two subsets X and Y such that each edge has one end in X and one end in Y.
\end{definition}

We can see that $G_1$ is a bipartite graph, defined above by \cite{bondymurty}. Thus, \(G_1\) is 2-colorable as all bipartite graphs are 2-colorable \cite{asratian1998bipartite}. This means \(\chi^{*}(G_1) = 2\). \(G_1\) is an example graph that satisfies our given conditions \((1)\) and \((2)\). The optimal coloring is shown in Figure \ref{fig:ex1-optimal}.

Although Heuristic \ref{heuristic-greedy} and Heuristic \ref{heuristic-welsh} do not always produce an optimal coloring for bipartite graphs, Heuristic \ref{heuristic-dsatur} always will produce an optimal coloring (2 colors). Heuristic \ref{heuristic-dsatur}, however, does not produce optimal colorings for all graphs. Both of these results are shown and proven in \cite{brelaz}. We chose not to create examples of these.

\begin{prop}
Heuristic \ref{heuristic-mis} also does not always produce an optimal solution, i.e. it does not always produce a minimum coloring of a graph.
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from earlier. We've constructed a graph \(G_2 = (V,E)\) for this example. This is shown Figure \ref{fig:ex2-uncolored}.

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{images/mis-1.png}
\caption{Uncolored original graph \(G_2\)}\label{fig:ex2-uncolored}
\end{figure}

By applying heuristic \ref{heuristic-mis} and finding maximal independent sets, we see that \(G_2\) is \emph{5-colorable}. This is shown Figure \ref{fig:ex2-colored}.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-2.png}
  \caption{Coloring of \(G_2\) after applying heuristic \ref{heuristic-mis}}\label{fig:ex2-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/mis-3.png}
  \caption{Optimal coloring of \(K\)}\label{fig:ex2-optimal}
  \end{figure}
\end{multicols}

This is a non-optimal solution. We can see that the subgraph created by vertices \(\{5, 6, 7, 8\}\) is complete and thus we need at least 4 colors. This graph is \emph{4-colorable} however, meaning \(\chi^{*}(K) = 4\). Thus, we can see that heuristic \ref{heuristic-mis} does not always give us an optimal solution. The optimal coloring of $G_2$ is shown in Figure \ref{fig:ex2-optimal}.

\begin{prop}
Heuristic \ref{heuristic-greedy} and heuristic \ref{heuristic-welsh} have an upper bound of \(\Delta + 1\).
\end{prop}

To show this, we will create another example that satisfies conditions \((1)\) and \((2)\) from above. After doing some research into bipartite graphs, we learned that \emph{crown graphs} are excellent at showing how bad greedy heuristics can be, as shown and defined in \cite{kordecki}.

\begin{definition}[Crown Graph]
A crown graph \(CR_n = (V, E)\) is an undirected graph such that \(U = \{u_1, u_2, \dots, u_i\}\) and \(W = \{w_1, w_2, \dots, w_j\}\) where \(V = U \cup W\), \(U \cap W = \emptyset\), and \(|U| = |W|\). There is an edge from \(u_i\) to \(w_j\) whenever \(i \neq j\). A crown graph can also be viewed as a complete bipartite graph where the edges of a perfect matching have been removed.
\end{definition}

We constructed a simple crown graph, \(G_3 = (V, E)\). We've let \(|V| = 10\) for this example.

\begin{figure}[H]
\centering
\includegraphics[scale=0.38]{images/graph-4.png}
\caption{Uncolored original graph \(G_3\)}
\end{figure}

We can see that \(\Delta(G_3) = 4\). We can also see \(d_{G_3}(v) = 4\) for all \(v \in V\). Thus, in both heuristics \ref{heuristic-greedy} and \ref{heuristic-welsh}, the algorithm would pick an order arbitrarily. We can show using this crown graph the worst-case scenario of these heuristics.

\begin{multicols}{2}
  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-5.png}
  \caption{Worst-case coloring of \(G_3\)}\label{fig:ex3-colored}
  \end{figure}

  \begin{figure}[H]
  \centering
  \includegraphics[scale=0.38]{images/graph-6.png}
  \caption{Optimal coloring of \(H\)}\label{fig:ex3-optimal}
  \end{figure}
\end{multicols}

\vspace{-0.3cm}

As shown in Figure \ref{fig:ex3-colored}, using either heuristic with this ordering, we get \(\chi(H) = 5\). This gives us \(\frac{|V|}{2}\) colors. This is the worst-case for a crown graph as shown in \cite{johnson}, but this graph also demonstrates the upper bounds for these heuristics.

Let's take a look at how this looks in general. Let \(P = (V, E)\) be a simple, complete graph. We can see that \(\chi(H) = \Delta(H) + 1\). This is very easy to see in a \emph{complete} graph, where all vertices are connected to every other vertex. This means all vertices have degree \(|V| - 1\). Thus, every time we color a node, a new color is needed. And since we have \(\Delta(P) = |V| - 1\) and \(|V|\) vertices, we will need \(\Delta(P) + 1\) colors. This is stated in \emph{Brooks' Theorem}.

\begin{theorem}[Brooks' Theorem]
For any connected undirected graph \(G\) with maximum degree \(\Delta\), the chromatic number of \(G\) is at most \(\Delta\) unless \(G\) is a complete graph or an odd cycle, in which case the chromatic number is \(\Delta + 1\).
\end{theorem}

The proof of Brook's Theorem can be found in \cite{lovasz}. Overall, heuristics can produce undesirable results compared to the optimal solution. In graph \(G_3\), in the worst case, some heuristics produce \(\chi(H) = 5\) when \(\chi^{*}(H) = 2\) as it is a bipartite graph. This is shown in Figure \ref{fig:ex3-optimal}.

\vspace{-0.1cm}

\section{Results}
Our main goal was to compare these heuristics to find out which heuristic was most effective in efficiently approximating the vertex coloring problem. we wanted to compare both quality (number of colors used) and efficiency (running time). For this study, we decided to omit the maximal independent set heuristic (heuristic \ref{heuristic-mis}) as it was so different from the other 3 as well as including another NP-hard problem. We implemented the three heuristics in Java and ran simulations according to the following:
\begin{itemize}
  \setlength\itemsep{0.4em}
  \item Graphs were generated based on the \emph{Erdős–Rényi} model \cite{erdds1959random}. This model generates a random graph with approximately $p$ percent of edges.
  \item We analyzed graphs with $10\%, 20\%, \ldots, 90\%$ edges. We called this the type of the graph. This gave us both sparse and dense graphs.
  \item Simulations were run on graphs with 5 different amount of vertices: 50, 100, 250, 500, 1000.
\end{itemize}

We used statistical analysis to compare the number of colors used and the running time of each heuristic based on edge percentage (type) and then also looked at how this changed as the number of vertices changed.

\begin{figure}[H]
  \includegraphics[width=\linewidth]{../poster/figures/plot-colors.pdf}
  \caption{Comparing number of colors used for the three heuristics}\label{fig:compare-colors}
\end{figure}

Based on computational data of the amount of colors used, as shown in Figure \ref{fig:compare-colors}, we found:
\begin{itemize}
  \item Heuristic \ref{heuristic-greedy} (Greedy) uses on average 2.52 colors more than Heuristic \ref{heuristic-welsh} (Welsh-Powell).
  \item Heuristic \ref{heuristic-welsh} (Welsh-Powell) uses on average 2.79 colors more than Heuristic \ref{heuristic-dsatur} (DSATUR).
\end{itemize}
This led us to the following partial order based on colors used, where $X \preceq Y$ means heuristic $X$ uses significantly more colors than heuristic $Y$.
\begin{center}
  \textbf{Greedy $\preceq$ Welsh-Powell $\preceq$ DSATUR}
\end{center}
As vertices increased, the partial order defined above stays consistent. Heuristic \ref{heuristic-dsatur} (DSATUR) always used the least colors while Heuristic \ref{heuristic-greedy} (Greedy) always used the most.

Based on computational data of the running time, as shown in Figure \ref{fig:compare-time}, we found:
\begin{itemize}
  \item Welsh-Powell is significantly slower than Greedy.
  \item DSATUR is significantly slower than Welsh-Powell.
\end{itemize}
This led us to the following partial order based on running time, where $X \preceq Y$ means heuristic $X$ uses significantly less time than heuristic $Y$.
\begin{center}
  \textbf{Greedy $\preceq$ Welsh-Powell $\preceq$ DSATUR}
\end{center}

As the edge percentage increases:
\begin{itemize}
  \item Heuristic \ref{heuristic-dsatur} (DSATUR) running time increases.
  \item Heuristic \ref{heuristic-greedy} (Greedy) and Heuristic \ref{heuristic-welsh} (Welsh-Powell) running time increases until 50\% edges then decreases until 90\% edges.
  \item The difference in running time between DSATUR and Greedy as well as DSATUR and Welsh-Powell grows larger and larger.
\end{itemize}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{../poster/figures/plot-time.pdf}
  \caption{Comparing running time of each heuristic for each edge percentage}\label{fig:compare-time}
\end{figure}

\section{Conclusion}
Although we were not able to compare the coloring of the heuristics to the optimal coloring, we were able to determine which heuristics gave the closest to the optimal solution. We found that Heuristic \ref{heuristic-dsatur} (DSATUR) produced the lowest number of colors. However, DSATUR also utilized the most time especially as the edge percentage of the graph grew. This is undesirable, because in large graphs, this defeats our purpose of using heuristics to efficiently approximate the VCP.

Based on our overall results, we would suggest using Heuristic \ref{heuristic-welsh} (Welsh-Powell) for most graphs. This is because the extra time needed from Heuristic \ref{heuristic-greedy} (Greedy) to Heuristic \ref{heuristic-welsh} (Welsh-Powell) to achieve better accuracy is not as significant compared to the extra time needed from Heuristic \ref{heuristic-welsh} (Welsh-Powell) to  Heuristic \ref{heuristic-dsatur} (DSATUR).
